---
title: "Reduced Model"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mgcv)
library(dplyr)
library(ggplot2)
library(ggnewscale)
library(survival)
library(patchwork)
library(purrr)
library(data.table)
library(ggnewscale)
library(mic.sim)
```
  
  
We simulate a data set with a heavily censored upper component. 
  
```{r}
set.seed(501)
n = 300
ncomp = 2
pi = function(t) {
  z <- 0.07 + 0.04 * t - 0.00045 * t^2
  tibble("1" = 1 - z, "2" = z)
}
`E[X|T,C]` = function(t, c)
{
  case_when(
    c == "1" ~ -4.0 + (0.24 * t) - (0.0055 * t^2),
    c == "2" ~ 3 + 0.001 * t,
    TRUE ~ NaN
  )
}
t_dist = function(n){runif(n, min = 0, max = 16)}
attr(t_dist, "min") = 0
attr(t_dist, "max") = 16
sd_vector = c("1" = 1, "2" = 1.05)
low_con = -3
high_con = 1
scale = "log"
example_data_reduced = simulate_mics(n = n, t_dist = t_dist, pi = pi, `E[X|T,C]` = `E[X|T,C]`, sd_vector = sd_vector, covariate_list = NULL, covariate_effect_vector = c(0), low_con = low_con, high_con = high_con, scale = "log") %>% suppressMessages()
```

The upper component barely includes any observations within the range of tested concentrations. It will be difficult to estimate mu for the upper component when we have very little information about the center of the distribution or its width.

```{r}
example_data_reduced %>% 
  mutate(verbose_comp = case_when(
    comp == 1 ~ "Component 1",
    TRUE ~ "Component 2"
  )) %>% 
  ggplot() +
  geom_segment(aes(x = t, xend = t, y = left_bound, yend = right_bound, color = verbose_comp), data = (. %>% filter(left_bound != -Inf & right_bound != Inf)), alpha = 0.3) +
  geom_segment(aes(x = t, xend = t, y = right_bound, yend = left_bound, color = verbose_comp), data = (. %>% filter(left_bound == -Inf) %>% mutate(left_bound = low_con - 2)), arrow = arrow(length = unit(0.03, "npc")), alpha = 0.3) +
  geom_segment(aes(x = t, xend = t, y = left_bound, yend = right_bound, color = verbose_comp), data = (. %>% filter(right_bound == Inf) %>% mutate(right_bound = high_con + 2)), arrow = arrow(length = unit(0.03, "npc")), alpha = 0.3) +
  geom_point(aes(x = t, y = left_bound, color = verbose_comp), data = . %>% filter(left_bound != -Inf), alpha = 0.3) +
  geom_point(aes(x = t, y = right_bound, color = verbose_comp), data = . %>% filter(right_bound != Inf), alpha = 0.3) +
  geom_function(fun = function(t){`E[X|T,C]`(t, c = "1")}, aes(color = "Component 1 Mean")) +
  geom_function(fun = function(t){`E[X|T,C]`(t, c = "2")}, aes(color = "Component 2 Mean")) +
  xlim(attr(t_dist, "min") ,attr(t_dist, "max")) +
  ylim(low_con - 2, high_con + 2) %>% suppressWarnings()
```

We attempt to fit the full model to the heavily censored data set and by looking at the cross validation results we find that none of the tested degree sets are able to converge with any consistency, suggesting this modeling approach is not appropriate for this data set.

```{r}
out_full = fit_EM(
  max_degree = 3, visible_data = example_data_reduced, verbose = 1, max_it = 30, nfolds = 5
)
```
```{r}
out_full$cv_results
```

To resolve this issue, we use the reduced model, which allows us to specify a component to treat as "fixed", which in this context means we assume it lies entirely above the tested concentration range, so that any observations within the tested range cannot belong to this component. By making this assumption, we are able to fit the remaining component and estimate the weights of the components over time. 

Note that observations above or below the tested concentrations can still be weighted as part of the remaining component that we are estimating the mean and width of.

```{r}
out_reduced = fit_EM(
  approach = "reduced", fixed_side = "RC", extra_row = FALSE, max_degree = 4, visible_data = example_data_reduced, verbose = 1, ecoff = 2
)
```


```{r}
plot_fm(out_reduced, ecoff = 2, add_log_reg = TRUE)
```







Next we simulate a data set similar to the last, except the upper component has slightly more overlap with the range of tested concentrations.

```{r}
set.seed(502)
n = 300
ncomp = 2
pi = function(t) {
  z <- 0.07 + 0.04 * t - 0.00045 * t^2
  tibble("1" = 1 - z, "2" = z)
}
`E[X|T,C]` = function(t, c)
{
  case_when(
    c == "1" ~ -4.0 + (0.2 * t) - (0.0055 * t^2),
    c == "2" ~ 2 + 0.001 * t,
    TRUE ~ NaN
  )
}
t_dist = function(n){runif(n, min = 0, max = 16)}
attr(t_dist, "min") = 0
attr(t_dist, "max") = 16
sd_vector = c("1" = 1, "2" = 1.05)
low_con = -3
high_con = 1
scale = "log"
example_data_reduced_2 = simulate_mics(n = n, t_dist = t_dist, pi = pi, `E[X|T,C]` = `E[X|T,C]`, sd_vector = sd_vector, covariate_list = NULL, covariate_effect_vector = c(0), low_con = low_con, high_con = high_con, scale = "log") %>% suppressMessages()
```

```{r}
example_data_reduced_2 %>% 
  mutate(verbose_comp = case_when(
    comp == 1 ~ "Component 1",
    TRUE ~ "Component 2"
  )) %>% 
  ggplot() +
  geom_segment(aes(x = t, xend = t, y = left_bound, yend = right_bound, color = verbose_comp), data = (. %>% filter(left_bound != -Inf & right_bound != Inf)), alpha = 0.3) +
  geom_segment(aes(x = t, xend = t, y = right_bound, yend = left_bound, color = verbose_comp), data = (. %>% filter(left_bound == -Inf) %>% mutate(left_bound = low_con - 2)), arrow = arrow(length = unit(0.03, "npc")), alpha = 0.3) +
  geom_segment(aes(x = t, xend = t, y = left_bound, yend = right_bound, color = verbose_comp), data = (. %>% filter(right_bound == Inf) %>% mutate(right_bound = high_con + 2)), arrow = arrow(length = unit(0.03, "npc")), alpha = 0.3) +
  geom_point(aes(x = t, y = left_bound, color = verbose_comp), data = . %>% filter(left_bound != -Inf), alpha = 0.3) +
  geom_point(aes(x = t, y = right_bound, color = verbose_comp), data = . %>% filter(right_bound != Inf), alpha = 0.3) +
  geom_function(fun = function(t){`E[X|T,C]`(t, c = "1")}, aes(color = "Component 1 Mean")) +
  geom_function(fun = function(t){`E[X|T,C]`(t, c = "2")}, aes(color = "Component 2 Mean")) +
  xlim(attr(t_dist, "min") ,attr(t_dist, "max")) +
  ylim(low_con - 2, high_con + 2) %>% suppressWarnings()
```

We can fit a reduced model this time, similar to the prior example, except we can use the `extra_row = TRUE` option to slightly change the assumption for the fixed side to be that the component lies entire above the range of tested concentrations OR at the highest tested concentration (in the plot, this looks like adding the top "row" of the graph to the set of observations that can be at least partially weighted as part of the fixed components). 
Note that extra_row = TRUE works similarly when the lower component is the one assumed to be fixed, where the set of observations that can be weighted as part of the fixed component are those lower than the lowest tested concentration and those between the lowest concentration and the second lowest.
```{r}
out_reduced_ex = fit_EM(
  approach = "reduced", fixed_side = "RC", extra_row = TRUE, max_degree = 4, visible_data = example_data_reduced_2, verbose = 1
)
```

```{r}
plot_fm(out_reduced_ex, title = "Extra Row")
```


We can contrast this to the same model but fitted with extra_row = FALSE. 

```{r}
out_reduced_no_ex = fit_EM(
  approach = "reduced", fixed_side = "RC", extra_row = FALSE, max_degree = 4, visible_data = example_data_reduced_2, verbose = 1
)
```



```{r}
plot_fm(out_reduced_no_ex, title = "No Extra Row")
```


